---
title: "Partial Least Squares Regression for Beta Regression Models"
author: "Frédéric Bertrand"
output: github_document
---

[![CRAN status](https://www.r-pkg.org/badges/version/plsRbeta)](https://cran.r-project.org/package=plsRbeta)

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi=300,fig.width=7,
  fig.keep="all"
)
```

# plsRbeta

The goal of plsRbeta is to provide Partial least squares Regression for (weighted) beta regression models (Bertrand 2013,  <http://journal-sfds.fr/article/view/215>) and k-fold cross-validation of such models using various criteria. It allows for missing data in the explanatory variables. Bootstrap confidence intervals constructions are also available.

## Installation

You can install the released version of plsRbeta from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("plsRbeta")
```

You can install the development version of plsRbeta from [github](https://github.com) with:

```{r, eval = FALSE}
devtools::install_github("fbertran/plsRbeta")
```

## Example

### Using a model matrix
Fit a plsRbeta model using a model matrix.

```{r exampleXY, cache=TRUE}
data("GasolineYield",package="betareg")
yGasolineYield <- GasolineYield$yield
XGasolineYield <- GasolineYield[,2:5]
library(plsRbeta)
modpls <- plsRbeta(yGasolineYield,XGasolineYield,nt=3,modele="pls-beta")
print(modpls)
```

Additionnal values can be retrieved from the fitted model.
```{r exampleXY2}
modpls$pp
modpls$Coeffs
modpls$Std.Coeffs
modpls$InfCrit
modpls$PredictY[1,]
rm("modpls")
```

###Formula support

Fit a plsRbeta model using formula support.
```{r exampleformula, cache=TRUE}
data("GasolineYield",package="betareg")
modpls <- plsRbeta(yield~.,data=GasolineYield,nt=3,modele="pls-beta", verbose=FALSE)
print(modpls)
```

Additionnal values can be retrieved from the fitted model.
```{r exampleformula2}
modpls$pp
modpls$Coeffs
modpls$Std.Coeffs
modpls$InfCrit
modpls$PredictY[1,]
```

###Information criteria and cross validation

```{r cv, cache=TRUE}
data("GasolineYield",package="betareg")
set.seed(1)
bbb <- PLS_beta_kfoldcv_formula(yield~.,data=GasolineYield,nt=3,modele="pls-beta",verbose=FALSE)
kfolds2CVinfos_beta(bbb)
```

###Bootstrap of the coefficients

Computing bootstrap distributions
```{r boot, warning=FALSE, cache=TRUE}
data("GasolineYield",package="betareg")
set.seed(1)
GazYield.boot <- bootplsbeta(modpls, sim="ordinary", stype="i", R=250)
```

Boxplots of the bootstrap distributions
```{r bootboxplots}
plsRglm::boxplots.bootpls(GazYield.boot)
```

Confidence intervals for the coefficients of the model based on the bootstrap distributions
```{r bootconfint, warning=FALSE}
plsRglm::confints.bootpls(GazYield.boot)
```

Plot of the confidence intervals for the coefficients of the model based on the bootstrap distributions
```{r bootplotconfint, warning=FALSE}
plsRglm::plots.confints.bootpls(plsRglm::confints.bootpls(GazYield.boot))
```
